The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015

### 2017-08-19 paper/paper.Rnw
Added 80/20-70/30 gaps in appendix - all working when restarted Rstudio

<<default_conf>>=
  library(matrixStats)
  library(knitr)
  library(arm)
  library(saves)
  library(haven)
  library(PISA2000lite)
  library(PISA2003lite)
  library(PISA2006lite)
  library(PISA2009lite)
  library(PISA2012lite)
  library(intsvy)
  library(cimentadaj) # devtools::install_github("cimentadaj/cimentadaj")
  library(countrycode) # For region variable
  library(car)
  library(SAScii)
  library(inequalityintsvy) # devtools::install_github("cimentadaj/inequalityintsvy")
  library(lme4)
  library(modelr)
  library(tidyverse)
  library(ggrepel)
  library(xtable)
  library(ggthemes)

@@

% Replace the Y_i with the chosen y from the previos equation.
\begin{equation}
\gamma_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{within country} \ j
\gamma_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{for each country} \ j
\end{equation}

@@

\begin{figure}
\begin{center}
<<graph_all_gaps>>=
complete_data_topbottom_80 <-
  pisa_preparer(results_math_80, results_read_80) %>%
  mutate(type = "80th/20th SES gap")

complete_gaps <-
  complete_data_topbottom %>%
  bind_rows(complete_data_topbottom_80)
  bind_rows(complete_data_topbottom_70)

complete_gaps %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  filter(type_test == "math",
         country %in% countries) %>%
  ggplot(aes(as.character(wave), difference, colour = type, linetype = type, group = type)) +
  geom_line(stat = "smooth", method = "lm", size = 0.7) +
  facet_wrap(~ country, ncol = 5) +
  scale_colour_discrete(name = "Type of gaps") +
  scale_linetype_discrete(name = "Type of gaps") +
  scale_y_continuous(name = "Gaps in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        panel.grid.major.y = element_line(colour = "grey"),
        legend.title = ) +
  ggtitle("Evolution of the different achievement gaps")
@
\caption{Evolution of the achievement gap for several gaps}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
<<percentage_change_8020>>=
perc_increase_fun(complete_data_topbottom_80) %>%
  perc_graph("math",
             "80/20 achievement gap",
             "Percentage change from 2000 to 2015")
@
\caption{Percentage change in the 80/20 achievement gap from 2000 to 2015}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
<<percentage_change_7030>>=
perc_increase_fun(complete_data_topbottom_70) %>%
  perc_graph("math",
             "70/30 achievement gap",
             "Percentage change from 2000 to 2015")
@
\caption{Percentage change in the 70/30 achievement gap from 2000 to 2015}
\end{center}
\end{figure}

@@

>>>>

<<default_conf>>=
  library(matrixStats)
  library(tidyverse)
  library(knitr)
  library(arm)
  library(saves)
  library(haven)
  library(PISA2000lite)
  library(PISA2003lite)
  library(PISA2006lite)
  library(PISA2009lite)
  library(PISA2012lite)
  library(intsvy)
  library(cimentadaj) # devtools::install_github("cimentadaj/cimentadaj")
  library(countrycode) # For region variable
  library(car)
  library(SAScii)
  library(inequalityintsvy) # devtools::install_github("cimentadaj/inequalityintsvy")
  library(lme4)
  library(modelr)
  library(ggrepel)
  library(xtable)
  library(ggthemes)

@@

% Replace the Y_i with the chosen y from the previos equation.
\begin{equation}
\gamma_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{for each country} \ j
\end{equation}

@@

@@

\pagebreak

write_rds(results_math_80, "./data/math_80")
write_rds(results_read_80, "./data/read_80")
write_rds(results_math_topmid_80, "./data/math_topmid_80")
write_rds(results_read_topmid_80, "./data/read_topmid_80")
write_rds(results_math_midbottom_80, "./data/math_midbottom_80")
write_rds(results_read_midbottom_80, "./data/read_midbottom_80")

write_rds(results_math_70, "./data/math_70")
write_rds(results_read_70, "./data/read_70")
write_rds(results_math_topmid_70, "./data/math_topmid_70")
write_rds(results_read_topmid_70, "./data/read_topmid_70")
write_rds(results_math_midbottom_70, "./data/math_midbottom_70")
write_rds(results_read_midbottom_70, "./data/read_midbottom_70")

<<eval = F>>=
results_math_80 <- read_rds("./paper/data/math_80.Rdata")
results_read_80 <- read_rds("./paper/data/read_80.Rdata")
results_math_topmid_80 <- read_rds("./paper/data/math_topmid_80.Rdata")
results_read_topmid_80 <- read_rds("./paper/data/read_topmid_80.Rdata")
results_math_midbottom_80 <- read_rds("./paper/data/math_midbottom_80.Rdata")
results_read_midbottom_80 <- read_rds("./paper/data/read_midbottom_80.Rdata")

results_math_70 <- read_rds("./paper/data/math_70.Rdata")
results_read_70 <- read_rds("./paper/data/read_70.Rdata")
results_math_topmid_70 <- read_rds("./paper/data/math_topmid_70.Rdata")
results_read_topmid_70 <- read_rds("./paper/data/read_topmid_70.Rdata")
results_math_midbottom_70 <- read_rds("./paper/data/math_midbottom_70.Rdata")
results_read_midbottom_70 <- read_rds("./paper/data/read_midbottom_70.Rdata")
@

\begin{figure}
\begin{center}
<<graph_all_gaps,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
complete_data_topbottom_80 <-
  pisa_preparer(results_math_80, results_read_80) %>%
  mutate(type = "80th/20th SES gap")

complete_gaps <-
  complete_data_topbottom %>%
  bind_rows(complete_data_topbottom_80) %>%
  bind_rows(complete_data_topbottom_70)

complete_gaps %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  filter(type_test == "math",
         country %in% countries) %>%
  ggplot(aes(as.character(wave), difference, colour = type, linetype = type, group = type)) +
  geom_line(stat = "smooth", method = "lm", size = 0.7) +
  facet_wrap(~ country, ncol = 5) +
  scale_colour_discrete(name = "Type of gaps") +
  scale_linetype_discrete(name = "Type of gaps") +
  scale_y_continuous(name = "Gaps in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        panel.grid.major.y = element_line(colour = "grey"),
        legend.position = "bottom") +
  ggtitle("Evolution of the different achievement gaps")
@
\caption{Evolution of the achievement gap for several gaps}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
<<percentage_change_8020,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
perc_increase_fun(complete_data_topbottom_80) %>%
  perc_graph("math",
             "80/20 achievement gap",
             "Percentage change from 2000 to 2015")
@
\caption{Percentage change in the 80/20 achievement gap from 2000 to 2015}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
<<percentage_change_7030,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
perc_increase_fun(complete_data_topbottom_70) %>%
  perc_graph("math",
             "70/30 achievement gap",
             "Percentage change from 2000 to 2015")
@
\caption{Percentage change in the 70/30 achievement gap from 2000 to 2015}
\end{center}
\end{figure}

### 2017-08-19 paper/paper.tex
Updated README with step-by-step of repo + deleted all useless scripts from before
Added 80/20-70/30 gaps in appendix - all working when restarted Rstudio

% This chunk needs to have cache = TRUE once you want to run it for like 2 hours until 
% the results_* models have been created once and the cache can save them

@@

The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}. This means that inequality starts strong before the schooling of a child begins and schools can do so much to close it. These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavourable conditions. But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\). Although this sounds logical, it's importance is not understood entirely.

@@

The first attempt to study the evolution of the achievement gap has found that the gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years. The literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence is emerging with a very similar landscape. The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940. This time-trend has allowed researchers to study achievement gaps in a lengthy period of time. Using this information, \citet{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising. Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citep{magnuson2008}. The gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).

@@

Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation. However likely, this is a highly contested topic. A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality. However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008}.

\footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.

@@

The reasons the authors find a reversal in the trend could be numerous and should be studied very closely. The authors incorporate a number of country-level indicators to try and explaind this change and suggest that the reversal is likely due to the high increase of preschool enrollment. They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially. Their conclusions are very speculative and have no empirical support, which is why this is still an open question.

@@

More formally, We're interested in studying
a) Which countries are experiencing changes in the achievement gap;
b) the rate at which each country-gap is widening/narrowing and;
c) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.
d) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;

@@

Yet one of the most relevant variables for our study is a composite SES index created by the PISA team. The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measured the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) \citep{oecd_glance_2002}. This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming all mentioned variables into comparable metrics across waves. To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.

@@

\begin{equation}
Y_i = \alpha + \beta_1 * AGE_i + \epsilon_i, \quad \epsilon_i ~ N(0, \sigma^2)
\end{equation}

for each wave,  where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (because all children should be 15 year olds) weighted by the student sample weights.

We then adjust \begin{math} \hat{Y_i} \end{math} by

% I think the denominator could be wrong! Because I include y_hat and that's what the equation is estimating
\begin{equation}
\hat{Y_i} = \frac{\hat{\epsilon_i}}{\sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
\end{equation}

@@

% Show graph of variance over time
% I'm showing standard deviations, is it better to show variance?

@@

In order to correct gap estimates for measurement error, I adjust student test scores(\begin{math} \hat{Y_i} \end{math}) by

\begin{equation}
\hat{Y_i} = \hat{Y_i} * \frac{1}{\sqrt{r}}
\end{equation}

where \begin{math}r\end{math} is the reliability score of the wave \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}. Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own. At the moment of writing this paper, they were unavailable. For this reason we implement the analysis following the original work of \citet{reardon2011}}. Each PISA survey provides a reliability indicator which we use accordingly. This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.

@@

\begin{equation}
Y_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{within country} \ j
\end{equation}

@@

We also calculate the standard error of this difference and use it to create uncertainty intervals in the estimation of time trends. We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size. Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible. For further exploration, we also compute the 90th/50th and 50th/10th SES gaps following the same method outlined above.

@@

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Wed Jul  5 17:51:02 2017
\scalebox{0.75}{
\begin{tabular}{lllrrlrr}
   \hline \
                          & & & Low SES & & & High SES \\
                          \cmidrule(l){3-5}
                          \cmidrule(l){6-8} Year & Countries & N & Avg score & S.E & N & Avg score & S.E \\ \hline2000 & Australia & 294 & 0.22 & 0.07 & 270 & 1.22 & 0.09 \\ 
  2000 & Austria & 294 & -0.04 & 0.11 & 235 & 1.15 & 0.15 \\ 
  2000 & Canada & 1970 & 0.30 & 0.05 & 1357 & 0.90 & 0.07 \\ 
  2000 & Denmark & 230 & -0.01 & 0.13 & 228 & 1.17 & 0.17 \\ 
  2000 & Finland & 261 & 0.42 & 0.12 & 267 & 0.89 & 0.15 \\ 
  2000 & France & 253 & 0.04 & 0.04 & 258 & 1.21 & 0.05 \\ 
  2000 & Germany & 250 & -0.64 & 0.04 & 303 & 1.81 & 0.05 \\ 
  2000 & Hungary & 291 & -0.46 & 0.09 & 254 & 1.66 & 0.12 \\ 
  2000 & Italy & 278 & -0.31 & 0.05 & 278 & 0.81 & 0.06 \\ 
  2000 & Netherlands & 129 & 0.46 & 0.08 & 158 & 1.25 & 0.11 \\ 
  2000 & Poland & 209 & -0.31 & 0.04 & 166 & 1.25 & 0.06 \\ 
  2000 & Spain & 318 & -0.27 & 0.05 & 340 & 1.09 & 0.07 \\ 
  2000 & Sweden & 241 & 0.12 & 0.10 & 246 & 1.06 & 0.13 \\ 
  2000 & United Kingdom & 531 & 0.16 & 0.04 & 415 & 1.33 & 0.06 \\ 
  2000 & United States & 222 & -0.43 & 0.02 & 164 & 1.74 & 0.03 \\ 
  2015 & Australia & 1694 & -0.15 & 0.04 & 1235 & 1.28 & 0.05 \\ 
  2015 & Austria & 676 & -0.30 & 0.07 & 705 & 1.53 & 0.10 \\ 
  2015 & Canada & 2215 & 0.20 & 0.03 & 1863 & 0.99 & 0.05 \\ 
  2015 & Denmark & 1013 & 0.09 & 0.08 & 595 & 1.08 & 0.10 \\ 
  2015 & Finland & 575 & 0.07 & 0.08 & 584 & 1.16 & 0.11 \\ 
  2015 & France & 570 & -0.37 & 0.02 & 615 & 1.66 & 0.03 \\ 
  2015 & Germany & 545 & -0.09 & 0.02 & 582 & 1.45 & 0.03 \\ 
  2015 & Hungary & 466 & -0.72 & 0.07 & 589 & 1.79 & 0.09 \\ 
  2015 & Italy & 947 & -0.20 & 0.03 & 1043 & 1.23 & 0.04 \\ 
  2015 & Netherlands & 521 & 0.13 & 0.04 & 525 & 1.24 & 0.06 \\ 
  2015 & Poland & 446 & 0.10 & 0.03 & 448 & 1.24 & 0.05 \\ 
  2015 & Spain & 608 & -0.25 & 0.03 & 703 & 1.24 & 0.04 \\ 
  2015 & Sweden & 527 & -0.20 & 0.06 & 542 & 1.32 & 0.09 \\ 
  2015 & United Kingdom & 1387 & -0.03 & 0.03 & 1388 & 1.15 & 0.04 \\ 
  2015 & United States & 585 & -0.39 & 0.01 & 544 & 1.22 & 0.01 \\ 
   \hline
\end{tabular}
}

@@

@@

@@

Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's an efficiency/equality tradeoff and if it stands over time. One particular topic that has been researched in the past is whether improving the overall cognitive development of a country compromises the distribution of the countries overall performance. For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance. This, however, can come at a price. If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance. We plot this relationship in figure 8.

\subsection{Efficiency-equality tradeoff}

@@

\caption{Time trends of achievement gap and average achiement between 2000 and 2015}
\end{center}
\end{figure}

@@

>>>>

% pisa_listcol.Rdata is a dataset produced by the source() expression from the above
% code chunk

% This chunk takes about 3-4 hours to run. Once I had them ready I saved them so you can read them
% as rds. If you run the 'compile pdf' this chunk is already cached (DONT CHANGE ANYTHING FROM
% THIS CHUNK, CAUSE OTHERWISE IT WILL RERUN FOR 3-4 HOURS). If you want to run the code without
% compiling sweave, comment the test_diff expressions and read as rds.

% Remember to uncomment when compiling sweave again, so the chunk remains intact.

@@

Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive and personality definition \citep{duyme1999, waldfogel1996}. Economists and educators have long neglected this branch of literature but the work of James Heckman has brought our attention to it. For example, he has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}. This means that inequality starts strong before the schooling of a child begins and schools have the potential to equalize the learning ground. These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavourable conditions. But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\). Although this sounds logical, it's importance is not understood entirely.

@@

The first attempt to study the evolution of the achievement gap has found that the gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years. The literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence is emerging with a very similar landscape. The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940. This time-trend has allowed researchers to study achievement gaps in a lengthy period of time. Using this information, \citet{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising. Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citep{magnuson2008}. The gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).
% widening over the years? Gosta, page 3
% income achievement gap, gosta page 3

@@

The first attempt to study the evolution of the achievement gap has found that the gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years. The literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence \citep{anna2016_global} is emerging with a very similar landscape. The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940. This time-trend has allowed researchers to study achievement gaps in a lengthy period of time. Using this information, \citet{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising. Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citep{magnuson2008}. The gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).

Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation. \citet{reardon2011} offers several possible links, with the most reasonable being that family investment patterns have changed so that high income families now invest more resources on their children. This explanation lies in the fact that increasing income became more strongly correlated with other positive family traits related to time allocation and welfare services.

However likely, this is a highly contested topic. A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality. However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008} \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970 where income inequality was increasing.}.
.

@@

The reasons the authors find a reversal in the trend could be numerous and should be studied very closely. The authors incorporate a number of country-level indicators to try and explaind this change and suggest that the reversal is likely due to the high increase of preschool enrollment. They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially. Their conclusions are very speculative and have no empirical support, which is why this is still an open question.
The reasons the authors find a reversal in the trend could be numerous and should be studied very closely. The authors incorporate a number of country-level indicators to try explain this change and suggest that the reversal is likely due to the high increase of preschool enrollment. In fact, the point out that in this same period (1998 - 2010) the SES gap in preschool enrollment decreased substantially. Their conclusions are very speculative and have no empirical support, which is why this is still an open question.

@@

More formally, we're interested in studying
a) Which countries are experiencing changes in the achievement gap;
b) the rate at which each country-gap is widening/narrowing and;
c) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.
d) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;

@@

Yet one of the most relevant variables for our study is a composite SES index created by the PISA team. The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measured the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) \citep{oecd_glance_2002}. This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming all mentioned variables into comparable metrics across waves.

The ESCS was derived from a principal component analysis of standardised variables, taking the factor scores for the first principal component as measures of the PISA index of economic, social and cultural status. All countries and economies (both OECD and partner countries/economies) contributed equally to the principal component analysis, while in previous cycles, the principal component analysis was based on OECD countries only. However, for the purpose of reporting the ESCS scale has been transformed with zero being the score of an average OECD student and one being the standard deviation across equally weighted OECD countries \citep{pisa_2015_results}.

To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.

@@

% Justify why you use the AGE variable
\begin{equation}
Y_i = \alpha + \beta_1 * AGE_i + \epsilon_i, \quad \epsilon_i ~ N(0, \sigma^2)
\end{equation}

for each wave, where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (following the same strategy as \citet{reardon2011}) weighted by the student sample weights.

We then adjust \begin{math} \hat{\gamma_i} \end{math} by

% I think the denominator could be wrong! Because I include y_hat and that's what the equation is estimating
\begin{equation}
\hat{\gamma_i} = \frac{\hat{\epsilon_i}}{\sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
\end{equation}

@@

% Try plotting this for each country and wave
% Like, every country as a row and 6 columns for each wave.
\begin{figure}
\begin{center}

@@

In order to correct gap estimates for measurement error, I adjust student test scores(\begin{math} \hat{\gamma_i} \end{math}) by

\begin{equation}
\hat{\gamma_i} = \hat{\gamma_i} * \frac{1}{\sqrt{r}}
\end{equation}

where \begin{math}r\end{math} is the reliability score of the wave \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}. The realibility estimates are calculated using Item Response Theory (IRT) analogues of traditional estimates of Person separation reliability such as internal consistency. Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own. At the moment of writing this paper, they were unavailable. For this reason we implement the analysis following the original work of \citet{reardon2011}}. Each PISA survey provides a reliability indicator which we use accordingly. This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.

@@

% Replace the Y_i with the chosen y from the previos equation.
\begin{equation}
\gamma_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{for each country} \ j
\end{equation}

% Try running also a linear model together with the multilevel model and ompcare results
% Brunno suggested this because countries which might have a very different SES estimate
% might be moved more towards the average.

@@

We also calculate the standard error of this difference and use it to create uncertainty intervals in the estimation of time trends. We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size. Given that including the SES \emph{dummy} reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible. For further exploration, we also compute the 90th/50th and 50th/10th SES gaps following the same method outlined above.
% Explain more in detail that it's a dummy and not three groups.

@@

% New table only has sample size, average score and s.e.
% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Wed Jul  5 17:51:02 2017
\scalebox{0.75}{
\begin{tabular}{lllrrlrr}
   \hline \
                          & & & Low SES & & & High SES \\
                          \cmidrule(l){3-5}
                          \cmidrule(l){6-8} Year & Countries & N & Avg score & S.E & N & Avg score & S.E \\ \hline2000 & Australia & 294 & 0.22 & 0.07 & 270 & 1.22 & 0.09 \\ 
  2000 & Austria & 294 & -0.04 & 0.11 & 235 & 1.15 & 0.15 \\ 
  2000 & Canada & 1970 & 0.30 & 0.05 & 1357 & 0.90 & 0.07 \\ 
  2000 & Denmark & 230 & -0.01 & 0.13 & 228 & 1.17 & 0.17 \\ 
  2000 & Finland & 261 & 0.42 & 0.12 & 267 & 0.89 & 0.15 \\ 
  2000 & France & 253 & 0.04 & 0.04 & 258 & 1.21 & 0.05 \\ 
  2000 & Germany & 250 & -0.64 & 0.04 & 303 & 1.81 & 0.05 \\ 
  2000 & Hungary & 291 & -0.46 & 0.09 & 254 & 1.66 & 0.12 \\ 
  2000 & Italy & 278 & -0.31 & 0.05 & 278 & 0.81 & 0.06 \\ 
  2000 & Netherlands & 129 & 0.46 & 0.08 & 158 & 1.25 & 0.11 \\ 
  2000 & Poland & 209 & -0.31 & 0.04 & 166 & 1.25 & 0.06 \\ 
  2000 & Spain & 318 & -0.27 & 0.05 & 340 & 1.09 & 0.07 \\ 
  2000 & Sweden & 241 & 0.12 & 0.10 & 246 & 1.06 & 0.13 \\ 
  2000 & United Kingdom & 531 & 0.16 & 0.04 & 415 & 1.33 & 0.06 \\ 
  2000 & United States & 222 & -0.43 & 0.02 & 164 & 1.74 & 0.03 \\ 
  2015 & Australia & 1694 & -0.15 & 0.04 & 1235 & 1.28 & 0.05 \\ 
  2015 & Austria & 676 & -0.30 & 0.07 & 705 & 1.53 & 0.10 \\ 
  2015 & Canada & 2215 & 0.20 & 0.03 & 1863 & 0.99 & 0.05 \\ 
  2015 & Denmark & 1013 & 0.09 & 0.08 & 595 & 1.08 & 0.10 \\ 
  2015 & Finland & 575 & 0.07 & 0.08 & 584 & 1.16 & 0.11 \\ 
  2015 & France & 570 & -0.37 & 0.02 & 615 & 1.66 & 0.03 \\ 
  2015 & Germany & 545 & -0.09 & 0.02 & 582 & 1.45 & 0.03 \\ 
  2015 & Hungary & 466 & -0.72 & 0.07 & 589 & 1.79 & 0.09 \\ 
  2015 & Italy & 947 & -0.20 & 0.03 & 1043 & 1.23 & 0.04 \\ 
  2015 & Netherlands & 521 & 0.13 & 0.04 & 525 & 1.24 & 0.06 \\ 
  2015 & Poland & 446 & 0.10 & 0.03 & 448 & 1.24 & 0.05 \\ 
  2015 & Spain & 608 & -0.25 & 0.03 & 703 & 1.24 & 0.04 \\ 
  2015 & Sweden & 527 & -0.20 & 0.06 & 542 & 1.32 & 0.09 \\ 
  2015 & United Kingdom & 1387 & -0.03 & 0.03 & 1388 & 1.15 & 0.04 \\ 
  2015 & United States & 585 & -0.39 & 0.01 & 544 & 1.22 & 0.01 \\ 
   \hline
\end{tabular}
}

@@

% Code chunks for the same plots as above for reading. Available upon request.

@@

% Add confidence intervals
\begin{figure}
\begin{center}

@@

Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's a tradeoff between the average country performance and the widening of the achievement gap and if it stands over time. One particular topic that has been researched in the past is whether improving the overall performance of a country compromises the equality of the educational system. For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance. This, however, can come at a price. If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance. We plot this relationship in figure 8.

\subsection{Performance-equality tradeoff}

@@

\caption{Time trends of achievement gap and average achievement between 2000 and 2015}
\end{center}
\end{figure}

% Add descriptive table with average test scores, degree of tracking, level of income inequality, among other things.

@@

\pagebreak

\section{Appendix}

{\centering \includegraphics[width=5.5in,height=5in]{figure/graph_all_gaps-1}

\caption{Evolution of the achievement gap for several gaps}
\end{center}
\end{figure}

{\centering \includegraphics[width=5.5in,height=5in]{figure/percentage_change_8020-1}

\caption{Percentage change in the 80/20 achievement gap from 2000 to 2015}
\end{center}
\end{figure}

{\centering \includegraphics[width=5.5in,height=5in]{figure/percentage_change_7030-1}

\caption{Percentage change in the 70/30 achievement gap from 2000 to 2015}
\end{center}
\end{figure}
